---
title: "Checking the calendar overlap of GTFS feeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking the calendar overlap of GTFS feeds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

Sys.setenv(OMP_THREAD_LIMIT = 2)

library(gtfstools)
```

Routing and accessibility analyses often combine several GTFS feeds, such as those of the different transit agencies of a metropolitan region. Routing engines such as `{r5r}` and OpenTripPlanner work on a given departure date, which must have service in every feed: otherwise, one of the networks silently drops out of the analysis. Feeds are published at different times and cover different periods, so such a date is not always easy to find.

`get_calendar_overlap()` returns the periods in which all the given feeds have service, and how many trips each feed runs on each day. It is based on the `check_gtfs_overlap()` function shared by [@higgicd](https://github.com/higgicd) in [issue #85](https://github.com/ipea/gtfstools/issues/85).

```{r, message = FALSE, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
```

# Finding a date with service in all feeds

Feeds can be given as paths to their `.zip` files. In this case, the large `shapes` and `stop_times` tables are not read. With `resolution = "periods"`, the function returns the periods in which all feeds have service:

```{r}
spo_path <- system.file("extdata/spo_gtfs.zip", package = "gtfstools")
poa_path <- system.file("extdata/poa_gtfs.zip", package = "gtfstools")

overlap <- get_calendar_overlap(c(spo_path, poa_path), resolution = "periods")
overlap
```

Each row is a period of consecutive days on which all feeds have service, with its first and last days (both included) and its length. The São Paulo and Porto Alegre feeds overlap for 91 days, from January 18 to April 18, 2019. Any of these days can be used, but a regular weekday away from holidays is usually the safest choice, e.g. the first Wednesday of the period (`format(x, "%u")` numbers weekdays from 1, Monday, regardless of the locale):

```{r}
period_days <- seq(overlap$start_date[1], overlap$end_date[1], by = "day")
departure_date <- period_days[format(period_days, "%u") == "3"][1]
departure_date
```

Feeds already read with `read_gtfs()` (and possibly edited afterwards) can be given as a list, whose names are used as feed names: `get_calendar_overlap(list(spo = spo_gtfs, poa = poa_gtfs), resolution = "periods")`.

A feed has service on a day when at least one of its services runs on it, according to the weekdays and dates in `calendar` and the exceptions in `calendar_dates`. Services not used by any trip are ignored.

# Visualising service periods

When the feeds don't overlap, the result has no rows:

```{r}
ber_path <- system.file("extdata/ber_gtfs.zip", package = "gtfstools")
feeds <- c(spo_path, poa_path, ber_path)

get_calendar_overlap(feeds, resolution = "periods")
```

`plot = TRUE` shows why. It returns a `{ggplot2}` timeline with one bar per feed, spanning its service days, and the periods in which all feeds have service shaded in the background. The result is a regular `ggplot` object, so it can be customised with `{ggplot2}` functions, e.g. to zoom in on a period, since the São Paulo feed spans more than 12 years:

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 2.5, dpi = 96}
get_calendar_overlap(feeds, resolution = "periods", plot = TRUE) +
  coord_cartesian(xlim = as.Date(c("2018-12-01", "2021-07-01")))
```

The Berlin feed starts in November 2020, after the São Paulo feed ends, so no date has service in all three feeds.

# Comparing service levels

The overlap does not guarantee a similar level of service in every feed: on some days, a feed may run only a few of its trips. The default, `resolution = "daily"`, returns the number of trips each feed runs on each of its service days, and whether all feeds have service on that day:

```{r}
daily_trips <- get_calendar_overlap(c(spo_path, poa_path))
head(daily_trips[daily_trips$overlap, ])
```

Trips listed in `frequencies` count once per departure. The São Paulo sample feed, for instance, describes all its service this way. With `plot = TRUE`, the daily trips of each feed are plotted in a panel of their own, which makes weekly patterns and holidays easy to spot. In Porto Alegre, there are far fewer trips on weekends:

```{r, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4, dpi = 96}
get_calendar_overlap(c(spo_path, poa_path), plot = TRUE) +
  coord_cartesian(xlim = as.Date(c("2019-01-01", "2019-05-01")))
```

A good departure date is thus a regular weekday on which every feed runs its usual number of trips.

Once a date is chosen, the feeds can be combined into a single one with `merge_gtfs()`, or reduced to the services of that date with the filtering functions (see the [filtering vignette](filtering.html)).
